### 2016-04-22 manuscript-body.tex
Copy parts of exported dissertation into manuscript body
But, I forgot to apply `--no-wrap` to the pandoc export, so now I have
to go fix that.

\section{Introduction}\label{introduction}

>>>>

\subsection{Introduction}\label{introduction-2}

Although imaginative and carefully designed, the Harvard course taught
that there is only one right way to approach the computer, a way that
emphasizes control through structure and planning. There are many
virtues to this computational approach (it certainly makes sense when
dividing the labor on a large programming project), but Lisa and Robin
have intellectual styles at war with it. Lisa says she has ``turned
herself into a different kind of person'' in order to perform, and Robin
says she has learned to ``fake it.'' Although both women are able to get
good grades in their programming course, they represent casualties of
this war. Both deny who they are in order to succeed. (Turkle \& Papert,
1991, p. 165)

Over 20 years ago Sherry Turkle and Seymour Papert called for
epistemological pluralism in computing education. Their argument --- in
my view ---had multiple parts:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Programming a computer involved a shift in how one thought about the
  nature knowledge from a propositional \emph{what is} to an imperative
  \emph{how to} (Abelson \& Sussman, 1996; Papert, 1980).
\item
  Historically, there was a tendency to treat computers and the act of
  programming them as an extension of the same Western logico-deductive
  knowledge tradition that fueled science.
\item
  Recent scholarship, often at the intersection of post-structuralism
  and social studies of science, had been troubling the idea that
  knowledge-work in science proceeded solely by way of logical
  deductions (Keller, 1983; Latour, 1987; Traweek, 1988)
\item
  Data from professionals suggested programming, too, didn't proceed
  solely by way of the structured-planning approach emergent from
  Western logico-deductive knowledge traditions. There were many ways of
  knowing and constructing knowledge in programming.
\item
  A viewpoint of \emph{epistemological pluralism} --- embracing multiple
  kinds and ways of knowing --- was informing the field of science
  studies but not, to a large degree, how education researchers thought
  about or taught computing.
\item
  The lack of support for diverse ways of knowing in computing
  classrooms demonstrably hurt students.
\end{enumerate}

To Turkle and Papert (1991), the result of this culturally-rooted
epistemological tension was a war. More specifically, it was a war where
students of computing were the casualties, and the aggressor was a
distributed failure to recognize, embrace, and support diverse ways of
knowing. While I think their strong-form characterization of a ``war''
no longer applies, their consideration of epistemological issues is
still relevant two decades later. An understanding that computing does
in fact involve diverse ways of knowing should continue to inform
research and discussions on learning

This paper tries to model what directs and sustains students'
in-the-moment activity when they program. Its focus is on early-stage
program design; analyzing how what students say, do, write, and gesture
\emph{even before they type code} can help us improve theories of
cognition and activity in learning programming. Using a ``revelatory
case study'' (Yin, 2009) of two students in an introductory programming
course, I argue the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Students' early-stage design activity reveals patterns outside the
  explanatory scope of misconception-based accounts of cognition.
\item
  For a subset of phenomena, we can recast students' productive
  capacities and their difficulties in terms of epistemological stances.
\item
  Such recastings are analytically powerful. Beyond Rebecca and Lionel,
  they could potentially explain the diversity of practices I saw other
  students take up.
\item
  As evidenced by work in science and math education, dynamic
  epistemological models can offer a lens for reforming assessment and
  instruction.
\end{enumerate}

\subsection{Literature Review}\label{literature-review}

I begin by exploring the historical legacy of Turkle and Papert's (1991)
remarks on epistemological pluralism in computing. I'll then describe
what I see as a well-intentioned obstacle: the research-based
preoccupation with students' misconceptions in computing education.
These two stra.

\subsubsection{Computing culture should support a diversity of ways of
knowing}\label{computing-culture-should-support-a-diversity-of-ways-of-knowing}

Turkle \& Papert (1991) describe a tension between individuals and a
kind of cultural collective. In this tension, individual students and
their personally-identified ways of knowing were in conflict with the
larger culture of computing and its manifested ways of knowing. Lisa,
for instance, was a Harvard student who identified as a poet:

Lisa experiences language as transparent, she knows where all the
elements are at every point in the development of her ideas. She wants
her relationship to computer language to be similarly transparent. When
she builds large programs she prefers to write her own smaller
``building block'' procedures even though she could use prepackaged ones
from a program library; she resents the latter's opacity. (Turkle \&
Papert, 1991, p. 164)

Turkle and Papert connect Lisa's programming practices to Lisa's
personal view of knowledge. She writes building block procedures because
of a continuity between herself as knower-of-her-own-poetry and that of
knower-of-her-own-programs. Across both contexts, Lisa \emph{resents
opacity} because it is an obstacle to her knowing. But, that resentment
slows her progress on projects in the class.

In part because of her commitments to transparency and the difficulty
they created, Lisa changed her approach to programming. She ultimately
had to, in her words, ``be a different kind of person with the machine''
(Turkle \& Papert, 1991, p. 164). Turkle and Papert explain:

She had been told that the ``right way'' to do things was to control a
program through planning and black­boxing, the technique that lets you
exploit opacity to plan something large without knowing in advance how
the details will be managed. Lisa recognized the value of these
techniques --- for someone else. She struggled against using them as the
starting points for her learning. Lisa ended up abandoning the fight,
doing things ``their way,'' and accepting the inevitable alienation from
her work. (Turkle \& Papert, 1991, p. 164)

The phrase ``inevitable alienation'' is particularly damning. One might
accuse Turkle and Papert (1991) of hyperbole, but research from other
disciplines suggests they're not overstating the case. Identity
alienation for epistemological reasons --- or some variation of them ---
is playing out in other subjects as well.

Connections and tensions between identities, learning, and personal ways
of knowing are not unique to computing education. Wortham (2006), for
example, detailed instances of social identification and academic
identity as \emph{jointly emerging} in a combined high school English
and History class. Nasir and colleagues explored how different learning
environments afford access to identity, domain practices, and
self-expression (Nasir \& Cooks, 2009; Nasir \& Hand, 2008). To single
out one study, Nasir \& Hand (2008) followed players ``from the
{[}basketball{]} court to the classroom.'' They found the court afforded
players a sense of role on the team, chances to be expressive through
play, and ``access to the domain {[}of basketball{]} as a whole'' (Nasir
\& Hand, 2006, p. 147). Comparatively, the mathematics classroom offered
little (if any) sense that students were on a team, a much narrower role
for them as having either right or wrong answers, and no strong sense of
self-expression or creativity through activity. These markedly different
contexts led not only to differences in kinds of activity but, Nasir and
Hand (2006) argue, divergent \emph{practice-linked identities} for the
players. In a sense, players had a different identity on the court than
they did in the mathematics classroom because who they were was strongly
coupled to a sense of their relationship with a practice.

Boaler's work on learning and identity in mathematics classrooms (Boaler
\& Greeno, 2000; Boaler, 1998, 2000, 2002) offers resonant findings. In
particular, Boaler (2000) strongly echoes the student voices in Turkle
and Papert (1991). Where Turkle and Papert's aggressor is the dominant
computing culture, Boaler describes a beast with similar effects in the
form of ``school mathematics'' --- the \emph{monotonous},
\emph{meaningless} (Boaler, 2002, pp. 383--384) imposition of routinized
mathematical procedures in ways that seem far removed from the real
world. Of school mathematics, Boaler writes:

School mathematics, for many of them, was \emph{of another world} and to
fully engage in that world, students needed to suspend their knowledge
of the real world, suppress their desire to interact with others, and
strive to reproduce standard procedures that held little meaning for
them. (Boaler, 2002, p. 392 emphasis in original)

Schoenfeld (1988, 1991) had already argued school mathematics could and
did depart substantially from the core of mathematical thinking. Boaler
established that such school mathematics could alienate learners
(Boaler, 1998, 2000) and influence their feelings about pursuing
mathematics after secondary school (Boaler \& Greeno, 2000).

Most recently, Danielak, Gupta, and Elby (in press) extended those
findings in engineering education. In their 3 ½ year case study of an
electrical engineering undergraduate named ``Michael,'' those authors
argued Michael's practice of sense-making in engineering coursework was
strongly coupled to his identity. Michael's passion for deeply
understanding concepts, much like Lisa's (Turkle \& Papert, 1991) desire
to understand every element of her program, did not mesh neatly with the
larger culture of his courses. But, because Michael's practice of
sense-making entwined with his identity, forces that pushed against
sense-making alienated him. That tension was particularly pointed when
Michael described his father's resistance to Michael's way of thinking:

{[}My dad said{]} ``you're just an undergraduate. Nobody expects
undergraduates to understand how anything works. That's why you go to
graduate school.'' I was like ``look, you know. I'm gonna be an unhappy
person if I have to....I have life goals other than to just get a good
grade on the exam. Other things are important to me.'' (Danielak et al.,
in press)

And, much like Turkle and Papert's (1991) Lisa, Michael learned to
curtail parts of his identity to get by in class:

I think the reason {[}pursuing deep learning{]} actually hasn't affected
my GPA is because I view learning as a hobby. So, as with any hobby, you
shouldn't let it interfere your GPA. But it \emph{is} one of my hobbies,
and I \emph{do} enjoy learning, I just---up to the point where I get my
grades done \{raises eyebrow\}. (Danielak et al., in press)

Ultimately, research suggests the identity-epistemology tensions Turkle
and Papert (1991) described in computing resonates with those found in
related disciplines. And, what such accounts --- particularly those of
Boaler and others (Boaler \& Greeno, 2000; Boaler, 1998, 2000, 2002;
Danielak et al., in press) --- have in common is a culturally-sanctioned
``right way'' (Turkle \& Papert, 1991, p. 164) of thinking or knowing
that alienates some students. Avowedly such ways of knowing have use and
value. Turkle \& Papert themselves acknowledge ``there are many
virtues'' to black-boxing and top-down design in programming (Turkle \&
Papert, 1991, p. 165). But,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  When a preponderance of results from science studies suggests a
  plurality of ways of knowing among practicing scientists and
  programmers, and
\item
  When research from education shows that the inflexible imposition of
  one way of knowing above all else is alienating students, then
\item
  It's worth reconsidering whether a narrow sense of what counts as
  knowing might still be hindering improvements in computing education.
\end{enumerate}

In the next section, I turn to misconceptions research in computing
education. In keeping with observations 1--3,

\subsubsection{Misconceptions research in computing education tends to
ignore students' productive
knowledge}\label{misconceptions-research-in-computing-education-tends-to-ignore-students-productive-knowledge}

I argue research
approaches that inflexibly privilege canonical knowledge do so at the
expense of other productive knowledge and ways of knowing that students
have.

In the past three decades, educational research has had a marked focus
on students' misconceptions in programming. It's a focus with a sensible
origin. Students get things wrong in programming --- often
systematically so --- and in ways that seem resistant to instruction.
The cause of those errors is theorized to be something cognitive,
whether it's a ``bug'' (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
rule'' (Fleury, 1991, 2000). Instruction should try to identify,
address, and correct these misconceptions (Clancy, 2004) because they
can represent barriers to learning.

How that line of thinking and research becomes problematic is two-fold.
First, when taken in total the alleged brokenness of student knowledge
begins eclipsing all else in studying the cognition of learning to
program. In other words, most cognitively-focused educational research
in computer science treats students as having varied degrees of
deficiency with respect to canonical knowledge. Below is an unordered,
partial sampling of topics about which researchers have documented
students' misconceptions. Note across the list the variation in both the
grain sizes of students' misconceptions and the programming languages in
which they manifest:

\begin{itemize}
\item
  Objects in object-oriented programming (Holland, Griffiths, \&
  Woodman, 1997)
\item
  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
  Paul \& Vahrenhold, 2013)
\item
  Programming statements in BASIC (Bayman \& Mayer, 1983)
\item
  Programming in Java (Fleury, 2000)
\item
  Programming in Pascal (Fleury, 1993)
\item
  Parameter-passing (Fleury, 1991)
\item
  Arrays in Java (Kaczmarczyk et al., 2010)
\item
  Objects in Java (Kaczmarczyk et al., 2010)
\item
  Algorithms and computational complexity (Trakhtenbrot, 2013)
\item
  Boolean logic (Herman et al., 2008)
\item
  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
\item
  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
\item
  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
\item
  The correctness of programs (Kolikant \& Mussai, 2008)
\end{itemize}

Clancy (2004) provides a comprehensive overview of this line of
research, though in the past decade it has only grown. Indeed, roughly
half the articles above were published in the ten years since Clancy's
overview.

Identifying and removing barriers to student learning seems like a good
thing. So, it should follow that cataloging student misconceptions and
developing remedies for them should also be a good thing. But, the
logical implication isn't that clean. In some cases students display
productive, useful knowledge that's either ignored or outright
criticized by researchers. Aligning students toward canonical knowledge
makes sense, but doing so at the expense of---or in direct contradiction
to---useful ways of knowing seems undesirable at best. Next, I expand on
two examples from misconceptions research in programming. Specifically,
I show how and why I think a misconceptions focus in programming casts
aside students' useful intuitions and understandings.

The first example comes from Kaczmarczyk et al. (2010). Part of that
study involved giving students snippets of Java code and asking students
to diagram (or pseudo-code) how the information would be stored in
memory. Below I have reproduced the code for Problem 2 (Kaczmarczyk et
al., 2010, p. 110):

Cheese{[}{]} cheeses = new Cheese{[}4{]};

Meat{[}{]} meats = new Meat{[}2{]};

Turkey turkey;

Ham ham;

RoastBeef roastBeef;

boolean lettuce = true;

boolean tomato = true;

SauceType sauceType = new SauceType();

int numMeat;

int numCheese;

In diagramming this information, a student in the study makes a mistake:

Student3 makes incorrect assumptions about connections between variables
to the extent that the student makes a mistake concerning the types of
the variables. As a result, the student places Objects of different
types in an array whose type matches none of them: ``And so because
there's two arrays, cheese and meats, uh, all those turkey and ham and
roast beef are gonna be sorted into the meats array.'' (Kaczmarczyk et
al., 2010, p. 110)

The researchers are correct in the sense that \textbf{turkey} and
\textbf{ham} and \textbf{roastBeef} will \emph{not} be sorted into the
\textbf{meats} array. First, there is no code here that places
\textbf{turkey} and \textbf{ham} in the array; there is only code that
declares them as variables. Moreover, as written, an attempt to place
\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} into the array
would fail. Because of type restrictions in Java, only objects of class
Meat (or objects that inherit from class Meat) can go in the array.
\textbf{turkey} and \textbf{ham} and \textbf{roastBeef} are, perhaps
confusingly, references to object instances of classes \textbf{Turkey}
and \textbf{Ham} and \textbf{RoastBeef}, so in the current snippet they
cannot enter an array of type \textbf{Meat} because (1) they don't yet
exist as objects and (2) even if they did exist, their types don't match
the array's type. The authors call this misconception \emph{semantics to
semantics}, which occurs ``when the student inappropriately
assume{[}s{]} details about the relationship and operation of code
samples, although such information was neither given nor implied''
(Kaczmarczyk et al., 2010, p. 110).

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon.\footnote{In this case, canon is
  the specifications and operations of the Java language and its
  compilers.} But, in their non-canonical thinking Student3 evidences
potentially productive insights about design. Precisely \emph{because}
there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
ham are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
into an array of type \textbf{Meat}. To do so, a designer could define
\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
from \textbf{Meat}.

Student3 has an idea about a relationship between entities where that
relationship is not specified in the code. Kaczmarczyk et al.
(Kaczmarczyk et al., 2010) focus only on the downside of the idea: the
student fails to display a proper understanding of how arrays work in
Java. Moreover, the student might be prone to similar mistakes of
inferring information that does not actually exist in code. But, there
is also an upside of this idea. Because Student3 is thinking about
real-world propositions like turkey and ham being kinds of meat, they
might be prepared to appreciate and discuss an object-oriented way to
put turkey in a \textbf{Meat} array. But, that possibility is
speculative conjecture. We can't know for certain whether Student3 could
be tipped into a productive object-oriented design activity around the
meats example because that question was not a focus of the research.

My second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's (1983) study of Pascal
programmers, part of which is discussed in Pea (1986). A student in
Bonar \& Soloway's study was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

Repeat

(1) Read a number (Num)

(1a) Count := Count + 1

(2) Add the number to Sum

(2a) Sum := Sum + Num

(3) until Count :=10

(4) Average := Sum div Num

(5) writeln (`average = `,Average)

The interviewer then asked whether (1a) and (2a) were ``the same kinds
of statements.'' That interchange is reproduced here:

Interviewer: Steps 1a and 2a: are those the same kinds of statements?

Subject: How's that, are they the same \emph{kind}. Ahhh, ummm, not
exactly, because with this {[}1a{]} you are adding - you initialize it
at zero and you're adding one to it {[}points to the right side of 1a{]}
which is just a constant kind of thing.

Interviewer: Yes

Subject: {[}points to 2a{]} Sum, initialized, to, uhh Sum to Sum plus
Num, ahh - thats {[}points to left side of 2a{]} storing two values in
one, two variables {[}points to Sum and Num on the right side of 2a{]}.
That's {[}now points to 1a{]} a counter, that's what keeps the whole
loop under control. Whereas, this thing {[}points to 2a{]} was probably
the most interesting thing\ldots{}about Pascal when I hit it. That you
could have the same, you sorta have the same thing here {[}points to
1a{]}, it was interesting that you cold have, you could save space by
having the Sum re-storing information on the left with two different
things there {[}points to right side of 2a{]}, so I didn't need to have
two. No, they're different to me.

Interviewer: So - in summary, how do you think of 1a?

Subject: I think of this {[}point to 1a{]} as just a constant, something
that keeps the loop under control. And this {[}points to 2a{]} has
something to do with something that you are gonna, that stores more
kinds of information that you are going to take out of the loop with
you. (Bonar \& Soloway, 1983, p. 12)

Pea's (1986) interpretation? ``Here, again, we see the student believing
that the programming language knows more about her intentions than it
possibly can'' (p. 32).

As before, this student has an idea about relationships in code. Pea
(1986) see the downside of her idea: believing PASCAL can understand
shades of programmer intent when, in fact, it cannot. And again, that
downside is real. It could cause trouble for this programmer later on if
she expects PASCAL to interpret her intent and it cannot.

In defense of the student, the question --- as asked --- is vague. Are
those statements the same \emph{to whom and in what way?} Pea (1986)
treats the data as though she meant ``the same to PASCAL.'' Indeed,
maybe she did, in which case his interpretation has traction. But,
another interpretation is that she meant to herself, or to someone else
reading the code. Those statements might not be the same \emph{to her}
because she treats (1a) as having a function of controlling iteration
while (2a)'s job is to combine two numbers into a new sum.

These two purposes, which for the sake of description I'll call
\emph{keeping control} and \emph{totaling up} are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

The structural aspects of a variable in a computer language are given
primarily by the rules for setting their values and for getting access
to their values. These rules apply in all contexts. In contrast, a
variable's functions might vary. Sometimes they might be described as
``a flag'' or more generally, as ``a communications device.'' At other
times a variable might function as ``a counter,'' ``data,'' or
``input.'' (diSessa, 1986, p. 202)

The student in Bonar \& Soloway's (1983) study did not show evidence of
understanding the structural similarities between (1a) and (2a) in her
pseudo-code. And, those authors as well as Pea (1986) justly insist that
similarity is important for students to understand. From a conceptual
standpoint, seeing the structural similarity constitutes a part of
``knowing'' PASCAL. But, even if knowing PASCAL were not the goal,
seeing the similarity helps one to take the perspective of a computing
agent that has no means for discerning programmer intent. Such
perspective-taking may help students avoid mistakes that arise from
over-assuming what a computer ``understands.''

The student did show evidence of understanding a functional difference
between (1a) and (2a), but Pea (1986) does not remark on that kind of
understanding at all.\footnote{Also glossed over is, to me, another
  important difference: a programmer might not know in advance which
  numbers are being passed in to the sum statement. So, in advance the
  programmer can say nothing about how the value of \textbf{Sum} will
  change as the loop iterates. In contrast, the programmer knows exactly
  how the value of Count will change with each loop iteration.} Again, I
claim this oversight is part of a subtle but observable trend in
programming misconceptions literature. While, or perhaps because
research has been so preoccupied warring with students' problematic
knowledge, it has sometimes failed to recover the productive knowledge
(or resources for building it) students have. In this example, the
student already has a grasp that syntactically similar statements could
serve different conceptual purposes. Hypothetically, the student might
use that information in design by calling up the ``◻=︎ + 1'' syntactical
template when the situation seems to demand \emph{keeping control},
while calling up ``◻= ︎ + number'' when \emph{totaling up} is the goal.
And, the idea that structurally identical symbol templates can serve
different functional and conceptual purposes fits precisely in line with
Sherin's (2001) theory of symbolic forms. For example, the symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while base±change describes a kind of accumulation.
Specifically, the terms in base±change ``play different roles'' (Sherin,
2001, p. 534) But, the two distinct conceptual schemata share what I
would argue is the same concrete symbol template of how to write an
expression: ◻=︎ + ◻.

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As I show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, I would argue
there are conceptual ways a programmer may need to think about
interpreting or writing a ◻=︎ +◻ symbol template that Sherin doesn't
enumerate.\footnote{One of the most obvious is the movement from seeing
  ``◻=︎ +◻'' as a statement of equality to seeing it as the storage of a
  sum to a variable.} To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

To return to misconceptions, what drives research on students'
misconceptions is largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
those are assuredly worthwhile goals. But, as I've argued, we can
already identify cases where a narrow misconceptions focus is silent
about or dismissive of students' useful intuitions. We can also identify
the further problems whereby unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need (Sherin, 2001) Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.\footnote{Worse still, in my view, is that there
  is a pattern of researchers failing to discuss or acknowledge
  students' productive knowledge \emph{even when their own data supports
  it}.} Such a perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge this way. I begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

\subsubsection{Not all cognitive programming bugs imply a problem with
the
student}\label{not-all-cognitive-programming-bugs-imply-a-problem-with-the-student}

What's curious about Pea's (1986) comments on Bonar \& Soloway (1983) is
that Pea drew different conclusions from the data than Bonar and Soloway
did. Pea (1986) emphasized that when the student thought two
semantically-equivalent assignment statements in PASCAL were different,
it was a problem of egocentrism: ``students assume that there is
\emph{more} of their meaning for what they want to accomplish in the
program than is actually present in the code they have written'' (p.
30). In other words, Pea treated the data as a fairly clear example of a
class of bugs. Specifically, he saw a bug class in which students simply
assumed the interpreter or runtime could infer the code author's shades
of intent.

Bonar and Soloway (1983), by contrast, were less quick to make
inferences either about the nature of the bug or about the intervention
it entailed. Rather than jump to definitive conclusions, they were
circumspect:

It is not clear exactly how to react to the bugs we have uncovered in
novice understanding of programming. In some cases it may be appropriate
to design new languages or constructs. Often, better instruction would
take care of the problem. The intent of our studies is to better
understand the source of the mismatches and misconceptions that cause
novice bugs. Only once a bug is uncovered and understood are we ready to
create a remedy for that bug. (Bonar \& Soloway, 1983, p. 12)

That the authors would even consider developing new constructs or
languages is a noteworthy distinction. Rather than assume in toto that
students with ``bugs'' had wrong knowledge, the authors instead suppose
cognitive bugs have plausible origins worth designing around. Moreover,
they treat students' divergence from canon as \emph{an opportunity for
research to learn from students}. Precisely because students saw
functional differences (cf., diSessa, 1986) in semantically-equivalent
PASCAL statements, Bonar and Soloway (1983) reflect that perhaps
programming languages should be more expressive:

We find it quite interesting that novices seem to understand the role or
strategy of statements more clearly than the standard semantics. Such
roles discussed here include ``counter variable,'' ``running total
variable,'' ``running total loop,'' and ``first, then rest loop''. (See
Soloway et al {[}1982b{]} for a detailed discussion of novice looping
strategies.) Much work in programming languages is concerned with
allowing a programmer to more accurately express his or her intentions
in the program. Perhaps we can learn something from novices here - our
programming systems should support recording the roles the programmer
intends for various statements and variables. (Bonar \& Soloway, 1983,
p. 12)

Again, what's noteworthy here is that rather than treating students'
non-canonical views as a burden for instruction, Bonar and Soloway
instead see them as an opportunity for programming language designers to
make languages better.

That insight---that instruction and design can meet novices where they
are---carries through to their final remarks about studying and
analyzing novice programming knowledge:

The experience and understanding of a novice are available for analysis.
In particular, our results suggest that the knowledge people bring from
natural language has a key effect on their early programming efforts.
Our work suggests that we need serious study of the knowledge novices
bring to a computing system. For most computerized tasks there is some
model that a novice will use in his or her first attempts. We need to
understand when it is appropriate to appeal to this model, and, when
necessary, how to move a novice to some more appropriate model. (Bonar
\& Soloway, 1983, p. 13)

I assume Bonar and Soloway structured their final line deliberately. If
so, their phrasing has three consequences:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Appealing to novice's existing models gets precedence. That is,
  understanding how to leverage novices' existing knowledge comes first
  in research.
\item
  Changing the models novices have comes next, and ``when necessary.''
\item
  They speak of ``how to move a novice to some more appropriate model,''
  which does not necessarily entail the removal or destruction of
  novice's existing models.
\end{enumerate}

Taken together, these points convey a sense of how Bonar and Soloway
view learning and instruction in computing. Instruction explicitly
includes appeals to prior models and knowledge students might already
have. Learning, meanwhile, involves the movement \emph{when necessary}
to more appropriate models of computation. As I explain in the next
section, such a view exactly aligns with a particular branch of
constructivism, where cognition is viewed as the complex activation of
manifold resources for thinking and knowing.

\subsubsection{Examples motivate the need for contextual-sensitivity in
modeling programming
cognition}\label{examples-motivate-the-need-for-contextual-sensitivity-in-modeling-programming-cognition}

Let's begin with two motivating examples. My aim with these examples is
to show how a practicing programmer might employ specific, distinct
conceptual models to reason locally about a piece of code. First,
consider variants of the PASCAL statements from Bonar and Soloway
(1983), where an assignment statement worked to increment a value and
store the result back to that value. In Table 2 below, I imagine five
different ways of writing a programming statement. Each of the five ways
is semantically equivalent to the others (or near enough for explanatory
purposes). And, in each example I add a layer of specificity to the
syntax. I also propose a corresponding interpretation of how I might
apply interpret and think about a given statement.\footnote{I assume for
  simplicity's sake that all variables and functions in my code samples
  evaluate to floating-point numbers that can be combined under
  addition. I also assume, crucially, that all statements are in the
  same hypothetical language and that the variable names connote nothing
  to the computational interpreter.}

\protect\hypertarget{ux5fToc252445956}{}{}Table -- Five different
interpretations of semantically equivalent programming statements in the
same language

\begin{center}
  \begin{tabular}{ | l | l |}
    1 & \texttt{x\ =\ x\ +\ d} \\
    2 & \texttt{x\ =\ x\ +\ update} \\
    3 & \texttt{x\_position\ =\ x\_position\ +\ update} \\
    4 & \texttt{x\_position\ =\ x\_position\ +\ update(t)} \\
    5 & \texttt{x\_position\ =\ x\_position\ +\ update(...)} \\
  \end{tabular}
\end{center}

% I stress that the third column is about \emph{how I might think about}
% each programming statement\emph{.} By no means am I making normative
% claims about how one \emph{ought} to think about it, or whether the
% statement actually does what its names might suggest it does. Rather,
% ``How I might think about it'' reflects the kind of local meaning or
% interpretation I might attach to such a statement when I work with it,
% given my understanding of its role and context.
%
% Consider specifically statements 1 and 4. Statement 1 could very well be
% an incrementer in some kind of iterative code. If I had to debug code
% that employs statement 1, what I might do is exploit my knowledge of
% what d is (does it hold a constant value? Is it 1?) and try inserting
% intermediate print statements into the iterative code. Doing so, I can
% inspect textually how values change with each iteration. In statement
% 4's case, it could very well be that the position-update code animates
% images on a screen. If there's a problem with that code, one of the
% easiest ways I might notice is that the acceleration seems off in the
% graphics. Consequently, my debugging might call upon knowledge I have
% from kinematics. I might try inserting code that draws a dot at the
% object's on-screen position with each iteration, leaving a trail I can
% visually inspect. I could then look at the path of the object's
% trajectory and the spacing patterns between successive dots as a
% first-pass test of whether my code achieves the motion I want.
%
% To be clear, it's not just that graphics might improve my efficiency in
% debugging a statement. Rather, applying an interpretive frame that
% treats an assignment statement as a kinematic position update lets me
% \emph{use conceptual knowledge from physics} to diagnose and fix
% problems in my code. If I view the assignment statement as saying
% something about the motion of an object (cf., Hammer, 1994, p. 165), a
% field of knowledge and concomitant techniques from physics becomes
% available to me to think with. But, if I focus only on the
% semantic-level equivalence of statements 1 through 5, there would be no
% obvious reason for me to access what I know about physics in order to
% reason about the code.
%
% My second example concerns statements that are syntactically-similar,
% rather than semantically-similar. I proposed that the statements in
% Table 2 all came from the same language. But, another phenomenon comes
% into play when different languages use the same symbology for
% conceptually different operations. Table 3 shows examples of what look
% like semantically-equivalent operations, but in fact are not.
%
% \protect\hypertarget{ux5fToc252445957}{}{}Table -- Three
% syntactically-similar statements with very different semantics
%
% \begin{supertabular}[]{@{}llll@{}}
% \toprule
% \textbf{Statement } & Programming Syntax & Language & How I think about
% it\tabularnewline
% \midrule
% \endhead
% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
% the string w\tabularnewline
% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
% layer of points onto a plot p\tabularnewline
% \bottomrule
% \end{supertabular}
%
% The catch here is that the plus operator takes on different roles in
% different languages because of how those languages define its use.
% Statement 1 increments a number in C; Statement 2 appends the letters
% ``ly'' to a string; Statement 3 adds a layer of points to a statistical
% graphics plot. These different kinds of operations become even more
% apparent and consequential when, for example, such statements are
% repeated. In the R/ggplot2 code below,\footnote{In R, the assignment
%   operator can be written as a directional arrow. The symbol
%   \textless{}- (``less-than, hyphen'') indicates the value on the right
%   side of the symbol is being assigned to the variable on the left side
%   of the symbol. As used, the symbol itself is semantically equivalent
%   to having written an equals sign (``=``).} I'm using multiple
% reassignment statements to compose a statistical graphics plot. The
% layered creation of a plot invites a very different kind of conceptual
% interpretation than, say, repeatedly accumulating numbers into a running
% sum:
%
% p \textless{}- InitializeGgplot\_1w()
%
% p \textless{}- p + GrandMeanLine(owp)
%
% p \textless{}- p + GrandMeanPoint(owp)
%
% p \textless{}- p + ScaleX\_1w(owp)
%
% p \textless{}- p + ScaleY\_1w(owp)
%
% p \textless{}- p + JitteredScoresByGroupContrast(owp, jj)
%
% One obvious reason for thinking about this code with a different
% interpretive frame is that numeric addition is commutative; composing a
% plot is not necessarily commutative.\footnote{To convince yourself that
%   plot composition isn't commutative, imagine a scale function that
%   squares up the aspect ratio of a plot and another scale function that
%   sets the aspect ratio to 1.5:1. Applying the square function last
%   produces a square plot; applying it first produces a rectangular plot.
%   String concatenation is also not necessarily commutative. ``cat'' +
%   ``dog'' evaluates to ``catdog,'' while ``dog'' + ``cat'' evaluates to
%   ``dogcat.''} So, despite the syntactic similarity, reassignments that
% compose a plot using reassignment (as above) does not obey the same
% rules as reassignments for a running total. But, even within this code
% block, statements that look alike perform operations of a different
% nature. While some expressions (e.g, ``+ GrandMeanLine(owp)'') add a
% visual layer to a plot, others modify features of the plot (e.g, ``+
% Scale\_X(owp)'', which adjusts the scales on the plot's x-axis to fit
% the numeric range of the data).
%
% Given these motivating examples, it seems sensible to think there's
% utility in a programmer having different conceptual metaphors available
% to think about and work with code. Example 1 shows that choosing to
% apply knowledge from physics to a piece of code can change the cognitive
% nature of debugging. There, debugging a position-update statement
% becomes, in part, reasoning kinematically about the properties of motion
% trails.\footnote{For comparison, consider the argument that ringing an
%   aircraft speedometer with physical markers changes the nature of
%   cognition when pilots work to land a plane (Hutchins, 1995b).} Example
% 2 shows that across languages, programmers might have to deploy
% different conceptual metaphors to reason about statements in a
% locally-consistent way. Knowing that plots in ggplot2 can be composed
% layer-by-layer with reassignment is crucial if you're trying to write or
% understand code that creates statistical graphics. But, I would argue
